Relational Parametricity for Algebraically Indexed Types

### 2012-06-21 paper/section-general-framework.tex
Tighten up the general framework section

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl{2},
  \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the index operations, we
  have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$ and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, with the following
  assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

We assume a countably infinite collection of index variable names $i,
i_1, i_2,$ and so on. \emph{Index contexts} $\Delta = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$ are lists of index expression
variable/sort pairs, such that all the index variable names are
distinct.

\begin{figure*}[t]
  \centering
  \textbf{Index contexts}\\

  $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types are constructed using the \TirName{TyExist}
rule.

@@

It will be technically convenient to express substitution of index
expressions in our framework in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,...,e_n)$ such that for each $1 \leq j \leq
n$, $\Delta \vdash e_j : s_j$. Given a simultaneous substitution
$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$.

It will be useful to think of any sequence of sorts as an index
context. In particular, we will make use of simultaneous substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$.

We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\Delta'$. The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. We will also define composition
of simultaneous substitutions below, but first we must define the
application of a simultaneous substitutions to index expressions.

Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined by the following two
clauses:
\begin{mathpar}
  \sigma^*i \isDefinedAs \sigma(i)

  \sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs \indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)
\end{mathpar}
Similarly, given a well-indexed type $\Delta' \vdash A \isType$, we
can apply $\sigma$ to $A$ to produce a new type $\sigma^*A$. The
application of a simultaneous substitution to a type is defined
by the following clauses:
\begin{mathpar}
  \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}

  \sigma^*\tyUnit \isDefinedAs \tyUnit

  \sigma^*(A \tyArr B) \isDefinedAs \sigma^*A \tyArr \sigma^*B

  \sigma^*(A \tyProduct B) \isDefinedAs \sigma^*A \tyProduct \sigma^*B

  \sigma^*(A + B) \isDefinedAs \sigma^*A + \sigma^*B

  \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A

  \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
\end{mathpar}
\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
  \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
    \isType$.
  \end{enumerate}
\end{lemma}

The \emph{composition} of two simultaneous substitutions $\Delta
\vdash \sigma \Rightarrow \Delta'$ and $\Delta' \vdash \sigma'
\Rightarrow \Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined
as $\Delta \vdash \sigma' \circ \sigma \isDefinedAs (\sigma^*e'_1,
..., \sigma^*e'_n) \Rightarrow \Delta''$.

Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index expressions is preserved by addition of extra
items to the context.

@@

Much of the power of indexing types by the expressions of an algebraic
theory comes from the equations of the theory. In the two-dimensional
geometry example, the types $\tyPrim{vec}{B,t_1 + t_2}$ and
$\tyPrim{vec}{B, t_2 + t_1}$ are considered equal by the type system,
due to the commutativity of the $+$ operation.

In the general framework, the equations between types are derived from
a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
e' : s$. We assume that all axioms $(\Delta \vdash e
\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted:
meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$.

Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
equality between index expressions $\Delta \vdash e \equiv e' : s$ by
the following rules, which allow us to use substitution instances of
axioms, and ensure that judgmental equality is a congruence relation
with respect to the index operations in $\IndexOpSet$:
\begin{mathpar}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}

  \inferrule*
  {\{\Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j\leq n}}
  {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
\end{mathpar}
We also assume the standard reflexivity, symmetry and transitivity
rules for the judgmental equality.

\begin{example}[Two-Dimensional Geometry]
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of orthgonal transformations is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
expressions generates the judgmental equality $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating judgmental equality on
types states that two applications of primitive types are equal if
their index expression arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for judgmental equality on types ensure that it
is a congruence relation on types, and that it is an equivalence
relation (i.e.~judgmental equality on types is reflexive, symmetric
and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
    \sigma^*e \equiv \sigma^*e' : s$; and
  \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
    \sigma^*A \equiv \sigma^* B \isType$.
  \end{enumerate}
\end{lemma}

A pair of simultaneous substitutions $\Delta \vdash \sigma, \sigma'
\Rightarrow \Delta'$ are defined to be judgmentally equal if their
component expressions are judgmentally equal in the context $\Delta$:
i.e.~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
simultaneous substitutions are judgmentally equal.

@@

Having defined the language of algebraically-indexed types, we now
turn to their denotational interpretation. In
\autoref{sec:index-erasure-semantics} we define an
\emph{index-erasure} interpretation of types that interprets every
well-indexed type as a set, ignoring the indexing expressions. Building on
the index-erasure semantics, we define in
\autoref{sec:relational-semantics} the \emph{relational}
interpretation of types.

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index expression equality: for
any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
substitutions $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash \sigma \equiv
\sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\
\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

@@

Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
the corresponding parts of \lemref{lem:tyeqsubst-erasure} to be
``well-typed''.

@@

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework.

@@

    \inferrule* [right=UnivAbs]
    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}

>>>>

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet =
  \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the
  index operations, we have additive group structure on
  $\SynTransl{2}$, multiplicative group structure on $\SynGL{1}$,
  $\SynGL{2}$ and $\SynOrth{2}$, injections from $\SynOrth{2}$ and
  $\SynGL{1}$ into $\SynGL{2}$, determinant, and absolute value. Thus,
  $\IndexOpSet = \{ 0, +, -, 1_G, -\cdot_G-, -^{-1_G}, \iota_O,
  \iota_1, \det, |\cdot| \}$, where $G \in \{\SynGL{1}, \SynGL{2},
  \SynOrth{2}\}$, with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

We assume a countably infinite collection of index variable names $i,
i_1, i_2,\ldots$. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
..., i_n \mathord: s_n$ are lists of variable/sort pairs, such that
all the variable names are distinct.

\begin{figure*}[t]
  \centering
  % \textbf{Index contexts}\\

  % $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

  % \bigskip

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types are constructed using the \TirName{TyExist}
rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use

@@

It will be technically convenient to express substitution of index
expressions in our framework in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,...,e_n)$ such that for each $1 \leq j \leq
n$, $\Delta \vdash e_j : s_j$. Given a simultaneous substitution
$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
the set of all simultaneous substitutions $\sigma$ such that $\Delta
\vdash \sigma \Rightarrow \Delta'$.

It will be useful to think of any sequence of sorts as an index
context. In particular, we will make use of simultaneous substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$. With
further abuse of notation, we write $\Delta \Rightarrow
\primTyArity(\tyPrimNm{X})$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$.

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index expressions is preserved by addition of extra
items to the context.

Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  Similarly, given a
well-indexed type $\Delta' \vdash A \isType$, we can apply $\sigma$ to
$A$ to produce a new well-indexed type $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}

The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of two
simultaneous substitutions $\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.

Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

@@

Much of the power of indexing types by the expressions of an algebraic
theory comes from the equations of the theory. In the two-dimensional
geometry example of \autoref{sec:motivating-examples}, the types
$\tyPrim{vec}{B,t_1 + t_2}$ and $\tyPrim{vec}{B, t_2 + t_1}$ are
considered equal by the type system, due to the commutativity of the
$+$ operation.

In the general framework, the equations between types are derived from
a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
e' : s$. We assume that all axioms $(\Delta \vdash e
\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted,
meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$
hold.

Given a set $\IndexAxiomSet$ of axioms, we generate the equality
judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
the following rules. The first allows us to use substitution instances
of axioms. The second ensures that equality is a congruence relation
with respect to the index operations in $\IndexOpSet$:
\begin{mathpar}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}

  \inferrule*
  {\{\Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j\leq n}}
  {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
\end{mathpar}
We also assume the standard reflexivity, symmetry and transitivity
rules for the equality judgment.

\begin{example}[Two-Dimensional Geometry]
  \fixme{Tie this example back to section 2}
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

The equality judgment $\Delta \vdash e \equiv e' : s$ on index
expressions generates the equality judgment $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating equality judgments on
types states that two applications of primitive types are equal if
their index expression arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for equality on types ensure that it is a
congruence relation on types, and that it is an equivalence relation
(i.e.,~equality on types is reflexive, symmetric and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

% FIXME: think about integrating this lemma into the running text
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
%     \sigma^*e \equiv \sigma^*e' : s$; and
%   \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
%     \sigma^*A \equiv \sigma^* B \isType$.
%   \end{enumerate}
% \end{lemma}

A pair of simultaneous substitutions $\Delta \vdash \sigma \Rightarrow
\Delta'$ and $\Delta \vdash \sigma' \Rightarrow \Delta'$ are defined
to be equal if their component expressions are equal in the context
$\Delta$: i.e.,~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all
$j$. We write $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
\Delta'$ when two simultaneous substitutions are equal.

@@

\fixme{Make sure that the semantics is motivated}
Having defined the language of algebraically-indexed types, we now
turn to their denotational interpretation. In
\autoref{sec:index-erasure-semantics} we define an
\emph{index-erasure} interpretation of types that interprets every
well-indexed type as a set, ignoring the indexing expressions. Building on
the index-erasure semantics, we define in
\autoref{sec:relational-semantics} the \emph{relational}
interpretation of types.

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types require an interpretation of index
contexts. We assign interpretations to index contexts in terms of sets
of \emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a (dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index expression equality: for
any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
substitutions $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

@@

Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

@@

We now present the rules for well-typed programs over the collection
of types we defined in \autoref{sec:algebraically-indexed-types}. Each
well-type program is assigned an index-erasure semantics, building on
the index-erasure semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus that instantiates our general framework.

@@

    \inferrule* [right=UnivAbs]
    {\Delta, i \mathord: s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}
